---
title: "Data for analysis with Rat Sightings"
author: "Jenesis Merriman"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Library calls
```{r}
library(tidyverse)
library(plotly)
library(readr)

knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 8, 
  fig.height = 6,
  out.width = "90%")

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis")

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

# With Rat Sightings Data

Load and clean NYC rat sighting data:
```{r rat data}
rat_df = read_csv("~/Desktop/p8105_final/Rat_Sightings.csv") %>%
  janitor::clean_names() %>%
  mutate(created_date = gsub(" .*","", created_date),
         address_type = str_to_title(address_type),
         city = str_to_title(city),
         borough = str_to_title(borough),
         location_type = recode(location_type, "Other (Explain Below)" = "Other")) %>%
  separate(created_date, into = c("month", "day", "year"), sep = "/") %>%
  mutate(year = as.numeric(year)) %>%
  filter(borough != "Unspecified") %>%
  select(month, day, year, location_type, address_type, city, borough, latitude, longitude)
```

Load NYC asthma emergency visit data for adults:
```{r asthma data}
asthma_df = read_csv("asthma_ER_visits_adults.csv") %>%
  janitor::clean_names() %>%
  rename(
    year = time
  ) %>%
  filter(geo_type == "Borough") %>%
  group_by(year, geography) %>%
  select(year, geography, age_adjusted_rate_per_10_000, estimated_annual_rate_per_10_000, number)
```

Merge datasets by year and borough: ???
```{r merge}
#this would need updating
asthma_rat_df <- left_join(rat_df, asthma_df, by = c('year' = 'year', 'borough' = 'geography'))
```

Map sightings by coordinate:
```{r map}
#turn it into a heat map since observation count
rat_map =
  rat_df %>% 
  plot_ly(
    lat = ~latitude, 
    lon = ~longitude, 
    type = "scattermapbox", 
    mode = "markers", 
    alpha = 0.2,
    color = ~borough) %>% 
  layout(
    mapbox = list(
      style = 'carto-positron',
      zoom = 9,
      center = list(lon = -73.9, lat = 40.7)),
    title = "<b> Map of Rat Sightings </b>",
      legend = list(title = list(text = "Borough", size = 9),
                    orientation = "v",
                   font = list(size = 9)))
rat_map
```

Heat map w/density??

```{r}
rat_df %>%
  plot_ly(
    type = 'densitymapbox',
    lat = ~latitude,
    lon = ~longitude,
    coloraxis = 'coloraxis',
    radius = 10,
    color = ~borough) %>%
  layout(
    mapbox = list(
      style = "stamen-terrain",
      zoom = 9,
      center = list(lon = -73.9, lat = 40.7)), 
    coloraxis = list(colorscale = "Viridis"))

```

Plot of sightings over time in NYC overall:

```{r line all}
overall_rat_line =
  rat_df %>%
  group_by(year) %>%
  count() %>%
  summarise(n_obs = n) %>% 
  ggplot(aes(x = year, y = n_obs)) + 
  geom_line() +
  labs(
    title = "Rat Sightings Over Time in NYC",
    x = "Year",
    y = "Number of Sightings")

overall_rat_line
```

Plot of sightings over time in NYC by borough:

```{r line borough}
rat_line = rat_df %>% 
  group_by(borough, year) %>%  
  count() %>%
  summarise(n_obs = n) %>% 
  ggplot(aes(x = year, y = n_obs , color = borough )) + 
  geom_line() +
  scale_x_continuous(breaks = seq(2010, 2022, by = 1)) +
  labs(
    title = "Rat Sightings Over Time by Borough",
    x = "Year",
    y = "Number of Sightings")

rat_line
```

Bar graph of all sightings by borough:

```{r bar borough}
rat_bar = 
  rat_df %>% 
  count(borough) %>% 
  mutate(borough = fct_reorder(borough, n)) %>% 
  ggplot(aes(x = borough, y = n, fill = borough)) + 
  geom_bar(stat = "identity") +
  labs(
    title = "Frequency of Rat Sightings by Borough (2010-2022)",
    x = "Borough",
    y = "Number of Sightings",
    fill = "Borough")
  
  #plot_ly(x = ~borough, y = ~n, color = ~borough, type = "bar", colors = "viridis")

rat_bar
```
Plot of frequency of sightings by borough and year:

```{r}
rat_bar_time =
  rat_df %>% 
  group_by(borough) %>%
  count(year) %>% 
  ggplot(aes(x = year, y = n, fill = borough)) +
  geom_bar(stat = "identity",
           position = "dodge") +
  scale_x_continuous(breaks = seq(2010, 2022, by = 1)) +
  labs(
    title = "Frequency of Sightings in Boroughs",
    x = "Year",
    y = "Number of Sightings",
    fill = "Borough")

rat_bar_time
```

Location type:
```{r}
location_bar =
  rat_df %>%
  count(location_type) %>%
   mutate(
    location_type = fct_reorder(location_type, n),
    ranking = min_rank(desc(n))) %>% 
  filter(ranking <= 10) %>% 
  arrange(n) %>%
  ggplot(aes(x = location_type, y = n, fill = location_type)) + 
  geom_bar(stat = "identity") +
  labs(title = "Top 10 Location Types",
       x = "Location Type",
       y = "Number of Sightings",
       fill = "location_type") + coord_flip() +
  theme(legend.position = "none") 

location_bar
```

```{r}
location_grid =
  rat_df %>%
  group_by(borough, location_type) %>%
  count() %>%
  summarise(n_obs = n) %>%
   mutate(
    location_type = fct_reorder(location_type, n_obs),
    ranking = min_rank(desc(n_obs))) %>% 
  filter(ranking <= 3) %>% 
  arrange(ranking) %>%
  ggplot(aes(x = location_type, y = n_obs, fill = location_type)) + 
  geom_bar(stat = "identity") +
  facet_grid(. ~ borough) +
  labs(title = "Top Sighting Locations by Borough",
       x = "Location Type",
       y = "Number of Sightings",
       fill = "location_type") +
  theme(axis.text.x = element_blank())
  
  
  #theme(axis.text.x = element_text(angle = 90, hjust = 1))

location_grid
```

Rat asthma plot:

```{r}
#idk yet i'm tired lol
```
